---
title: "Covariate-driven UK / GAM / RF and the bias-on-roads question"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariate-driven UK / GAM / RF and the bias-on-roads question}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 5)
```

Step 2 of the project showed that universal kriging and a thin-plate
GAM are statistically indistinguishable under spatial cross-validation,
but both produce a systematic bias of roughly -4 µg/m³ when applied to
held-out road-side monitors after training on the background network.

This vignette asks whether **adding covariates** (or **combining
stations with a `station_type` factor**) closes that bias.

```{r packages, message = FALSE}
library(aqsurface)
library(sf)
library(dplyr)
library(ggplot2)
```

## Sample data with `Road_Dist` + `DEM`

```{r data}
data("pm10_jan_s1",      package = "aqsurface")
data("pm10_jan_s1_road", package = "aqsurface")
data("stations_demo",    package = "aqsurface")

meta <- sf::st_drop_geometry(stations_demo)[, c("Station", "Road_Dist", "DEM")]

attach_meta <- function(df) {
  out <- merge(df, meta, by.x = "Station.ID", by.y = "Station")
  sf::st_as_sf(out, coords = c("X", "Y"), crs = 5181, remove = FALSE)
}

train   <- attach_meta(pm10_jan_s1)
holdout <- attach_meta(pm10_jan_s1_road)
targets <- decade_columns("pm10", month = 1, decade = "S1")
```

## Three scenarios

```{r scenarios}
common <- list(train = train, holdout = holdout, targets = targets,
               algorithms = c("uk", "gam", "rf"),
               cv_folds = 5, seed = 1, progress = FALSE)

baseline    <- do.call(benchmark_methods, common)
with_cov    <- do.call(benchmark_methods,
                       c(common, list(covariates = c("Road_Dist", "DEM"))))

# Combined training: Fixed + Road, with station_type as a factor covariate.
combined <- combine_stations(train, holdout)
combined_res <- benchmark_methods(
  train      = combined,
  holdout    = NULL,                            # everyone is in train
  targets    = targets,
  algorithms = c("uk", "gam", "rf"),
  covariates = "station_type",
  cv_folds   = 5, seed = 1, progress = FALSE
)

results <- dplyr::bind_rows(
  dplyr::mutate(baseline,     scenario = "baseline"),
  dplyr::mutate(with_cov,     scenario = "Road_Dist + DEM"),
  dplyr::mutate(combined_res, scenario = "combined + station_type")
) |>
  dplyr::filter(target == "ALL")

results |>
  dplyr::select(scenario, algorithm, strategy, n, rmse, bias, r)
```

## Reading the table

The legacy bias of about -4 µg/m³ should drop sharply when we add
`station_type` to a combined-station training set, because the model
can now learn an explicit road-vs-background offset rather than
having to extrapolate it. RF is expected to gain the most because it
can interact `station_type` with location non-linearly.

```{r plot}
results |>
  ggplot(aes(x = strategy, y = rmse, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  facet_wrap(~ algorithm) +
  labs(x = NULL, y = "RMSE (μg/m³)",
       title = "PM10 Jan S1: covariates and RF vs the baseline") +
  theme_bw()
```

## Caveats

* The RF in-sample RMSE is, like UK in-sample, biased downward
  because trees see every training row in some sub-sample.
* The combined-station scenario has no `held-out` strategy because
  road monitors are now in the training set; the spatial-cv result
  is the cleanest summary number.
* Real-world deployment would also want random k-fold CV across
  stations with at least 10-20 repeats; we use a single seed for
  this vignette to keep the runtime small.
