---
title: "Benchmarking UK and GAM with three validation strategies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking UK and GAM with three validation strategies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 5)
```

The legacy analysis evaluated UK and GAM with **in-sample** RMSE only,
which understates true predictive error. This vignette runs three
complementary validation strategies in one call.

```{r packages, message = FALSE}
library(aqsurface)
library(sf)
library(dplyr)
library(ggplot2)
```

## Data

```{r data}
data("pm10_jan_s1",      package = "aqsurface")
data("pm10_jan_s1_road", package = "aqsurface")

train <- sf::st_as_sf(pm10_jan_s1, coords = c("X", "Y"),
                      crs = 5181, remove = FALSE)
holdout <- sf::st_as_sf(pm10_jan_s1_road, coords = c("X", "Y"),
                        crs = 5181, remove = FALSE)
targets <- decade_columns("pm10", month = 1, decade = "S1")
```

## Run the benchmark

`benchmark_methods()` runs UK and GAM through three strategies and
returns one tidy tibble.

```{r run}
results <- benchmark_methods(
  train      = train,
  holdout    = holdout,
  targets    = targets,
  algorithms = c("uk", "gam"),
  cv_folds   = 5,
  fold_method = "kmeans",
  seed       = 1,
  progress   = FALSE
)

results |>
  dplyr::filter(target == "ALL") |>
  dplyr::select(algorithm, strategy, n, rmse, mae, bias, r)
```

## Visualise

```{r plot, fig.height = 4}
results |>
  dplyr::filter(target == "ALL") |>
  ggplot(aes(x = strategy, y = rmse, fill = algorithm)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(x = NULL, y = "RMSE (μg/m³)",
       title = "PM10 Jan S1: UK vs GAM across three validation regimes") +
  theme_bw()
```

## Interpretation

Three quantities show up cleanly here:

1. **In-sample RMSE** is the smallest for both methods, as expected.
   This is what the legacy scripts reported.
2. **Spatial CV RMSE** sits between in-sample and held-out. It
   quantifies pure spatial extrapolation error within the same
   monitor population (background-to-background).
3. **Held-out RMSE** on road-side monitors is the largest, with a
   negative bias, because the training set is exclusively
   background and systematically lower than road-side
   concentrations.

The gap between **spatial CV** and **held-out** is the part of the
error that any single-population model cannot fix. Step 3 of the
project will close that gap by adding a road-class covariate to
both UK (external drift) and GAM (`by`-factor smooth).
