---
title: "Comparing UK and GAM for PM10 in Seoul, January S1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing UK and GAM for PM10 in Seoul, January S1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 6)
```

```{r packages, message = FALSE}
library(aqsurface)
library(sf)
library(terra)
library(dplyr)
```

## Two ways to load the data

`aqsurface` ships with a small Jan S1 sample (objects `pm10_jan_s1`,
`pm10_jan_s1_road`, `stations_demo`, `ratio_demo`), so the vignette
knits without any external file. If the `AQSURFACE_DATA_DIR`
environment variable is set, the vignette switches to the full archive
instead.

```{r data}
have_full_data <- !is.na(aqs_data_dir(required = FALSE))

if (have_full_data) {
  data_dir <- aqs_data_dir()
  stations <- make_station_sf(file.path(data_dir, "stations_10km.shp"))
  parts    <- split_stations(stations)
  train_pm10   <- load_pollutant(file.path(data_dir, "pm10.RData"),
                                 season = "winter", station_type = "bk",
                                 stations = parts$train)
  holdout_pm10 <- load_pollutant(file.path(data_dir, "pm10.RData"),
                                 season = "winter", station_type = "rd",
                                 stations = parts$holdout)
} else {
  data("stations_demo",       package = "aqsurface")
  data("pm10_jan_s1",         package = "aqsurface")
  data("pm10_jan_s1_road",    package = "aqsurface")
  parts        <- split_stations(stations_demo)
  train_pm10   <- sf::st_as_sf(pm10_jan_s1,      coords = c("X", "Y"),
                               crs = 5181, remove = FALSE)
  holdout_pm10 <- sf::st_as_sf(pm10_jan_s1_road, coords = c("X", "Y"),
                               crs = 5181, remove = FALSE)
}

targets <- decade_columns("pm10", month = 1, decade = "S1")
grid    <- build_prediction_grid(train_pm10, n_x = 100, n_y = 100)
```

## Universal kriging

```{r uk}
uk <- predict_surface_batch(
  data      = train_pm10,
  targets   = targets,
  newgrid   = grid,
  algorithm = "uk",
  trend     = "const",
  cutoff    = 30000,
  width     = 3000,
  model     = "Ste",
  progress  = FALSE
)

plot_variogram_grid(uk$models, ncol = 5)
plot_surface_facet(uk$pred, limits = c(0, 150), ncol = 5)
```

## GAM

```{r gam}
gam <- predict_surface_batch(
  data      = train_pm10,
  targets   = targets,
  newgrid   = grid,
  algorithm = "gam",
  smooth    = "tp",
  method    = "REML",
  progress  = FALSE
)

plot_surface_facet(gam$pred, limits = c(0, 150), ncol = 5)
```

## Validation against road-side monitors

```{r validation}
uk_long  <- extract_at_stations(uk$pred,  holdout_pm10, targets)
gam_long <- extract_at_stations(gam$pred, holdout_pm10, targets)

dplyr::bind_rows(
  dplyr::mutate(compute_metrics(uk_long),  method = "UK"),
  dplyr::mutate(compute_metrics(gam_long), method = "GAM")
)
```

## Road-side post-processing (legacy comparison)

The legacy workflow multiplies background interpolations by an empirical
back-to-road ratio on road pixels. The package keeps this as a separate
post-processing step rather than folding it into the model.

```{r road, eval = FALSE}
# Requires the road raster, only available in AQSURFACE_DATA_DIR.
if (have_full_data) {
  ratio_tbl   <- tidy_ratio_table(
    read_pollutant_rdata(file.path(data_dir, "pm10.RData"))$win_ratio,
    targets = targets
  )
  road_raster <- terra::rast(file.path(data_dir, "..", "road_10km_re.tif"))
  uk_with_road <- apply_road_ratio(uk$pred, road_raster, ratio_tbl)
}
```
